---
title: "HW 5:  Factor and figure management; Repo hygiene"
author: "Sara Sheikholeslami"  
date: "`r format(Sys.Date())`"
output: 
  github_document
---

## Table of Contents

1. [Factor Management](#1)
    
    1.1. [Activity #1: Drop Oceania](#2)
    
    1.2. [Activity #2: Reorder the levels of country or continent](#3)
    
    1.3. [Activity #3: Practice writing to file and reading back](#4)
        
2. [File I/O](#5)

    2.1. [write_csv()/read_csv()](#6)
    
    2.2. [saveRDS()/readRDS()](#7)

3. [Visualization design](#8)

4. [Writing figures to file](#9)

5. [Clean up your repo!](#10)

6. [But I want to do more!](#11)

7. [Report your process](#11)

In this homework I will work with the gapminder data. Let's start by loading the required liberaries:
```{r}
suppressWarnings(suppressMessages(library(tidyverse)))
library(tidyverse)
library(gapminder)
library(tidyr)
library(plyr)
library(dplyr)
library(kableExtra)
library(knitr)
library(forcats)
```
Time to tackle the activities!

## <a id="1"></a> 1. Factor Management

**<a id="2"></a> 1.1. Activity #1: Drop Oceania**

Let's take a look at the data:

```{r}
str(gapminder)
```
We can see that the original gapminder dataset has 142 different countries, and 5 different continents. Lets learn more about Oceania by using 'fct_drop()' to only include Oceania:

```{r}
OC <- gapminder %>% 
  filter(continent %in% c("Oceania")) 

OC$country %>% 
  fct_drop() %>% 
  levels()
```
We can further see that there are two different countries under Oceania. To exclude Oceania, we can filter the data, and then use 'droplevels()':

```{r}
#filter the data and drop the extra level (Oceania)
gm_MinusOc <-gapminder %>% 
  filter(continent %in% c("Asia", "Africa", "Europe", "Americas"))%>%
  droplevels()

# let's take a look at the new dataframe
str(gm_MinusOc)
```


The new dataframe gm_MinusOc has only 4 continents, and the 2 countries under Oceania are dropped from country variable, leaving only 140 from the original 142 countries. Lets double check this with 'fct_count()'.

```{r}
fct_count(gm_MinusOc$continent)
```

**<a id="3"></a> 1.2. Activity #2: Reorder the levels of country or continent**

**<a id="4"></a> 1.3. Activity #3: Practice writing to file and reading back**
  
## <a id="5"></a> 2. File I/O

**<a id="6"></a> 2.1. write_csv()/read_csv()**

**<a id="7"></a> 2.2. saveRDS()/readRDS()**

## <a id="8"></a> 3. Visualization design

## <a id="9"></a> 4. Writing figures to file

## <a id="10"></a> 5. Clean up your repo!

## <a id="11"></a> 6. But I want to do more!


3. [Report your process](#11)


**<a id="2"></a> 1.1. Activity #2**

Make a tibble with one row per year and columns for life expectancy for two or more countries.
Use knitr::kable() to make this table look pretty in your rendered homework.
Take advantage of this new data shape to scatterplot life expectancy for one country against that of another.

Let's create a new data shape that looks at life expectancy for two or more countries per year:
```{r}
subset = gapminder %>%
  filter(country %in% c("Iran", "China")) %>%
  select(year, country, lifeExp) %>%
  spread(country, lifeExp)
```

Now let's make a tibble representing this new data shape:
```{r}
library("kableExtra")
html_table_width <- function(kable_output, width){
  width_html <- paste0(paste0('<col width="', width, '">'), collapse = "\n")
  sub("<table>", paste0("<table>\n", width_html), kable_output)}

knitr::kable(subset, format = "html", booktabs = TRUE, caption = 'Table of life expectancy changing over time on two different countries') %>%
  html_table_width(c(0,100,100))
```
Now let's plot the results:
```{r}
subset%>%
  ggplot(aes(x=China,y=Iran)) + geom_point()+labs(title=" life expectancy changing over time on two different countries")+geom_text(aes(label=year),hjust=1, vjust=0)+
   geom_smooth(se=FALSE, colour="red")

```

Let's try a different representation:
        
```{r}
subset = gapminder %>%
        select(year, country, lifeExp) %>%
        filter(country %in% c("Iran", "China"))

subset%>%
  ggplot(aes(x = year, y = lifeExp, colour = country)) +
  geom_point() + 
  geom_line() +
  theme(legend.title = element_text(size=12, face="bold"), legend.text = element_text(size=12)) +
  labs(x = "Year", y = "Life Expectancy", title = "life expectancy changing over time on two different countrie") 
```
        
        
**<a id="3"></a> 1.2. Activity #3**

Considering all the measures of life expectancy:
```{r}
subset2<-gapminder%>%
  group_by(continent,year)%>%
  summarise(meanLE=mean(lifeExp),medianLE=median(lifeExp),minLE=min(lifeExp),maxLE=max(lifeExp))

html_table_width <- function(kable_output, width){
  width_html <- paste0(paste0('<col width="', width, '">'), collapse = "\n")
  sub("<table>", paste0("<table>\n", width_html), kable_output)}

knitr::kable(subset2, format = "html", booktabs = TRUE, caption = 'Different measures of life expectancy (mean, median, min and max) for all possible combinations of continent and year') %>%
  html_table_width(c(0,100,100,100,100,100))
```




## <a id="11"></a> 3. Report your process

Overall I found this assignment easy to do. 
- I spent a lot of time trying to do 3d scatter plots but I was not able to generate a sensible visualization.... 
- I found using the join functions over and over a bit tedious!
