---
title: "HW 3: Use dplyr to manipulate and explore data (also use ggplot2)"
author: "Sara Sheikholeslami"  
date: "`r format(Sys.Date())`"
output: 
  github_document
---

# Overview
The following readme document is provided for hw03.
This README file includes:

1. [Your mission, high-level](#1)

2. [Task menu](#2)

3. [Companion graphs](#3)

4. [But I want to do more!](#4)

5. [Report your process](#5)

## <a id="1"></a> 1. Your mission, high-level

- `dplyr` will be used for data manipulation 
- `ggplot2` will be used for visualization 

## <a id="2"></a> 2. Task menu

Let's start by loading the required liberaries:

```{r}
suppressWarnings(suppressMessages(library(tidyverse)))
library(tidyverse)
library(gapminder)
```

***Task 1:*** Get the maximum and minimum of GDP per capita for all continents.

```{r}

knitr::kable(gapminder%>% 
  group_by(continent) %>% 
  dplyr::select(continent, gdpPercap) %>%
  summarise(maxGdp = max(gdpPercap), minGdp = min(gdpPercap)), align = 'c', format = 'markdown')
  
```
```{r}
ggplot(gapminder,aes(x=continent,y=gdpPercap))+geom_point(aes(color=continent))+theme_bw()+labs(title="continent vs gdpPercap")

```
Well the figure above is not really informative!

Let's only plot the maximum and minimum gdpPercap for all continents:

```{r}

gapminder%>% 
  group_by(continent) %>%
  mutate(minGdp = (min(gdpPercap)==gdpPercap)) %>% 
  mutate(maxGdp = (max(gdpPercap)==gdpPercap)) %>%
  #mutate(minMaxGdp = as.vector(rbind(min(gdpPercap),min(gdpPercap)))) %>% 
  #mutate(minMaxGdp = c(min(gdpPercap), max(gdpPercap))) %>%
  ggplot(aes(x=lifeExp , y=gdpPercap)) + geom_point(aes(color=minGdp , size=maxGdp)) +facet_wrap(~continent)
  
```

the different colored point indicates the minimum, and the big sized point indicates the maximum for each continent.

Alternatively:
```{r}
gapminder%>%
  group_by(continent)%>%
  mutate(minGdp=min(gdpPercap))%>%
  ggplot(aes(x=continent , y=minGdp)) + geom_point(aes(color=minGdp))


gapminder%>%
  group_by(continent)%>%
  mutate(maxGdp=max(gdpPercap))%>%
  ggplot(aes(x=continent , y=maxGdp)) + geom_point(aes(color=maxGdp))

```
This way we are only looking at the minimum and maximum GDP values for each continent!

***Task 2:*** Look at the spread of GDP per capita within the continents.

```{r}
knitr::kable(gapminder%>% 
  group_by(continent) %>% 
  dplyr::select(continent, gdpPercap) %>%
  summarise(maxGdp = max(gdpPercap), minGdp = min(gdpPercap), range =  IQR(gdpPercap)), align = 'c', format = 'markdown')
```

Let's start with histogram:

```{r}

set.seed(1)
mydf <- data.frame(continent = gapminder$continent, 
                   gdpPercap = gapminder$gdpPercap)
head(mydf)
tail(mydf)

library(lattice)
histogram(~gdpPercap | continent, data = mydf)
```

And a density plot:
```{r}
densityplot(~gdpPercap | continent, data = mydf)
```


Another way to look at spread is to use boxplots:
```{r}
ggplot(gapminder,aes(x=continent,y=gdpPercap)) + 
  geom_boxplot(outlier.colour = "light pink", show.legend = TRUE) + 
  geom_jitter(position = position_jitter(width = 1), alpha = 0.08)
```

and fitting a distribution:

```{r}

library(fitdistrplus)
library(logspline)
descdist(gapminder$gdpPercap, discrete = FALSE)

fit.weibull <- fitdist(gapminder$gdpPercap, "weibull")
plot(fit.weibull)

```
The data ranges from min:  241.1659 and  max:  113523.1.

***Task 3:*** Compute a trimmed mean of life expectancy for different years. Or a weighted mean, weighting by population. Just try something other than the plain vanilla mean.

**trimmed mean**
```{r}
library("kableExtra")
html_table_width <- function(kable_output, width){
  width_html <- paste0(paste0('<col width="', width, '">'), collapse = "\n")
  sub("<table>", paste0("<table>\n", width_html), kable_output)
}

knitr::kable(
subset<-gapminder%>%
  group_by(year)%>%
  summarise(meanT=mean(lifeExp,trim=0.5),meanW=weighted.mean(lifeExp,pop)), format = "html", booktabs = TRUE, caption = 'Trimmed mean and weighted mean of life expectancy for different years.') %>%
  html_table_width(c(50,100,100))
```

```{r}
ggplot(subset,aes(x = year, y = meanT))+ geom_point(color="blue")+
  geom_point(aes(x = year, y = meanW),color="red") +
  labs(x="year", y="Mean Life Expectancy", title="Mean Life Expectancy vs. year")
```


***Task 4:*** How is life expectancy changing over time on different continents?
```{r}
html_table_width <- function(kable_output, width){
  width_html <- paste0(paste0('<col width="', width, '">'), collapse = "\n")
  sub("<table>", paste0("<table>\n", width_html), kable_output)
}

subset<-gapminder%>% 
  group_by(year, continent)%>% 
  #group_by(year)%>% 
  summarise( meanT=mean(lifeExp,trim=0.5), meanW=weighted.mean(lifeExp,pop))
  #mutate(diff_lifeExp=c(0,diff(mean_lifeExp)))
knitr::kable(subset, format = "html", booktabs = TRUE, caption = 'Table of life expectancy changing over time on different continents') %>%
  html_table_width(c(100,100,100,100))
```
```{r}

gapminder%>%
  group_by(continent, year, lifeExp)%>% 
  summarise( meanT=mean(lifeExp,trim=0.5), meanW=weighted.mean(lifeExp,pop)) %>% 
ggplot(aes(year, lifeExp)) +
    facet_wrap(~ continent) +
    geom_point(alpha=0.3) +
  geom_smooth(aes(x = year, y = meanW),color="blue") +
  labs(x="year", y="Mean Life Expectancy", title="Mean Life Expectancy vs. year")
          
```

## <a id="3"></a> 3. Companion graphs

Please refer to previous section!

## <a id="4"></a> 4. But I want to do more!

No time... :(

## <a id="8"></a> 8. Report your process

Overall I found this assignment easy to do except for visualizing ***Task 4:***. It took me a LONG time to figure out how to plot the data in a sensible manner!
