---
title: "HW 1 Gapminder"
author: "Sara Sheikholeslami"  
date: "`r format(Sys.Date())`"
output: 
  github_document:
  toc: true
  #toc_depth: 4
---

```{r}
suppressWarnings(suppressMessages(library(tidyverse)))
library(tidyverse);
library(gapminder);

```

## 1.0 Add R Markdown and Markdown for Gapminder

## Overview
### [1.1 Basic info on a data frame](#1)

###  [1.2 `dplyr` fundamentals](#2)
#### [1.2.1 filter](#3)
#### [1.2.2 select](#4)
#### [1.2.3 pipeline](#5)
### [1.3  Including Plots](#6)

## <a id="1"></a> 1.1 Basic info on a data frame
### Number of columns:
```{r}
ncol(gapminder)
```

### Number of rows:
```{r}
nrow(gapminder)
```

## <a id="2"></a> 1.2 `dplyr` fundamentals

### <a id="3"></a> 1.2.1 filter: 
subsets data frames according to some logical expression. 

__Exercises__: Find all Europian and African countries, occuring in the '80s with a lifeExp ranging from 70 to 80. 
```{r}
filter(gapminder, continent %in% c("Europe", "Africa") & 
       year %in% 1980:1989 & lifeExp <= 75.0 &  lifeExp >= 65.0)
```

### <a id="4"></a> 1.2.2 select:
subsets data by columns/variable names.
__Exercises__: Take all entries of Europe and Africa occuring in the '80s with a lifeExp ranging from 70 to 80, and select the `country`, `year`, and `lifeExp` columns.

```{r}
select(filter(gapminder, continent %in% c("Europe", "Africa") & 
       year %in% 1980:1989 & lifeExp <= 75.0 &  lifeExp >= 65.0),
       country, year, lifeExp)
```

### <a id="5"></a> 1.2.3 pipeline:
When doing more than one operation, the __pipe__ operator `%>%` feeds the output of a function into another function

__Exercises__: Take all entries of Europe and Africa occuring in the '80s with a lifeExp ranging from 70 to 80, and select the `country`, `year`, and `lifeExp` columns. (similar to previous case)
```{r}
gapminder %>% 
    filter(continent %in% c("Europe", "Africa") & 
       year %in% 1980:1989 & lifeExp <= 75.0 & lifeExp >= 65.0) %>%
  select(country, year, lifeExp)
```

## <a id="6"></a> 1.3 Including Plots

Embeded plotes using only base R graphics:

```{r}
plot(lifeExp ~ continent, gapminder)
```

```{r}
hist(gapminder$lifeExp)
```
